---
title: "Distinctive Collexeme Analysis in Learner Language"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Reading in the data and prepare it for use

```{r}
library(rjson)
library(tidyverse)
library(rlist)

# topics where conditionals might be found
cond_topic_ids <- c(59, 104, 97, 109, 119, 82, 120, 100, 46, 88, 75, 108, 67, 106, 77, 68, 73, 107, 110, 74) %>%
  sort()

# read in CSV file with Conditionals Type II (CII) obtained from the analyses with Stanza (see Python scripts 01_ParseEssays.py and 02_IDConditionals.ipynb)
cond_ii <- read.csv2("./Python output files/02_conditionals_ii.csv", sep=",")

# read in the word counts per topic
topic_nwords <- fromJSON(paste(readLines("./Python output files/topic_word_count.json"), collapse="")) %>%
  enframe(name = "topic", value = "n_words")
topic_nwords$n_words <- as.numeric(topic_nwords$n_words)

# some topics have no occurrences of CII, but are still part of the analyzed subcorpus. Avoid overlooking them
cond_ii$topic <- factor(cond_ii$topic, levels = cond_topic_ids)

```

## Distinctive Collexeme Analysis

Instead of comparing the differences in collexeme attraction to a slot in two different constructions, we compare the difference between use of the collexeme in one structure at two different levels of proficiency. 


```{r}

# Lemmas with too few occurrences in the corpus might be unreliable. Only graph those with at least this many occurrences
MIN_LEMMA_OCCUR <- 5

# cross-tabulate as per Gries (2023 p. 361)
freqs_per_group <- table(lemma=cond_ii$cor_if_head_lemma, group=cond_ii$group) 

# Extract lemmas that appear often enough in the corpus
lemmas_threshold <- freqs_per_group %>%
  as.data.frame() %>%
  group_by(lemma) %>%
  summarise(
    sum = sum(Freq)
  )%>%
  filter(sum >= MIN_LEMMA_OCCUR) %>%
  pull(lemma) %>%
  as.character()

print(paste("Lemmas with at least", MIN_LEMMA_OCCUR, "occurences:", str_flatten_comma(lemmas_threshold), "(", length(lemmas_threshold), ")"))

# Go back to table format to be able to do Chi-square test
freqs_per_group <- freqs_per_group %>%
  as.data.frame() %>%
  filter(lemma %in% as.character(lemmas_threshold)) %>%
  xtabs(Freq ~ lemma + group, data = .)

# Chi-square test as per Gries (2023) ---> Obtain residuals
chisq <- chisq.test(freqs_per_group, correct = FALSE)
chisq_res <- chisq$residuals


freqs_with_resids <- data.frame(lower=freqs_per_group[,"lower"],
                        upper=freqs_per_group[,"upper"],
                        res_lower = chisq_res[, "lower"],
                        res_upper = chisq_res[, "upper"]) %>%
                        arrange(res_upper) %>%
                        drop_na()

print("Verbs disproportionally favored by learners at lower levels")
print(head(freqs_with_resids, 5))

print("Verbs disproportionally favored by learners at upper levels")
tail(freqs_with_resids, 5) %>%
  arrange(desc(res_upper)) %>%
  print()

```

### Visualizing verbal collexeme attraction to apodosis slot in upper vs. lower levels 

```{r}
# order by top upper residuals:
top_upper_res <- freqs_with_resids %>%
  arrange(desc(res_upper)) %>%
  filter(row.names(freqs_with_resids) %in% lemmas_threshold) # only graph the lemmas with enough occurrences

library(ggplot2)
top <- top_upper_res %>%
  mutate(lemma = row.names(.))

ggplot(top, aes(x = reorder(lemma, res_upper), y = res_upper, fill = res_upper > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "#4393C3", "FALSE" = "#D6604D"),
    labels = c("TRUE" = "Upper levels (B2–C1)", "FALSE" = "Lower levels (A2–B1)"),
    name   = "Level"
  ) +
  ylab("DCA Chi-squared residuals") +  
  xlab("Verb collexemes in CII apodosis")
```

## Formulaic to productive: lexical variety in apodosis verb, looking at Entropy
```{r}
# Calculating H
p_lower <- freqs_per_group[, "lower"] / sum(freqs_per_group[, "lower"])
p_upper <- freqs_per_group[, "upper"] / sum(freqs_per_group[, "upper"])

# remove 0's to be able to log
p_lower <- p_lower[p_lower > 0]
p_upper <- p_upper[p_upper > 0]

H_lower <- -sum(p_lower * log2(p_lower))
H_upper <- -sum(p_upper * log2(p_upper))

# relative entropy
Hrel_lower <- H_lower/ log2(length(p_lower))
Hrel_upper <- H_upper /log2(length(p_upper))


print(paste("Normalized entropy lower levels:", Hrel_lower))
print(paste("Normalized entropy higher levels:", Hrel_upper))
```



## Calculating dispersion (DP and DP_norm)

```{r}
per_topic_freqs <- cond_ii %>%
  group_by(topic) %>%
  count(topic, .drop = FALSE, name = "n_cii")

# add topic word count to frequency dataframe
per_topic_freqs$nwords_topic <- topic_nwords$n_words[match(topic_nwords$topic, per_topic_freqs$topic)]

total_words <- sum(per_topic_freqs$nwords_topic)

# total number of type II conditionals
total_n_cii <- sum(per_topic_freqs$n_cii)

# % of words in the subcorpus that each part ("topic") comprises.
per_topic_freqs$topic_perc_words <- per_topic_freqs$nwords_topic / total_words
 
# if a part has 10% of the words in the corpus, the, assuming a uniform distribution, it should have 10% of the conditionals
per_topic_freqs$expec_perc_ncii <- per_topic_freqs$topic_perc_words
# actual percentage of CII found in this part of the corpus
per_topic_freqs$obs_perc_ncii <- per_topic_freqs$n_cii / total_n_cii
# calculate absolute difference between observed and expected percentages
per_topic_freqs$abs_diff <- abs(per_topic_freqs$obs_perc_ncii - per_topic_freqs$expec_perc_ncii)

# Calculate DP
per_topic_dp <- sum(per_topic_freqs$abs_diff) / 2

# DP_norm according to Correction to Stefan Th. Gries’ “Dispersions and adjusted frequencies in corpora”, International Journal of Corpus Linguistics, 13:4 (2008), 403–437
per_topic_dpnorm <- per_topic_dp / (1 - 1/min(per_topic_freqs$nwords_topic))

print(paste("DP:", per_topic_dp))
print(paste("DP_norm:", per_topic_dpnorm))
```

